---
title: 'Practical 03 SG: LD and Haplotype estimation'
author: "Carlos Moyano & Kleber Reyes"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
editor_options:
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
rm(list = ls(all.names = TRUE))

```


```{r load packages, echo=FALSE, message=FALSE, warning=FALSE}
library(rstudioapi)
library(genetics)
library(HardyWeinberg)
library(LDheatmap)
library(haplo.stats)
set.seed(2209)
current_path <- getActiveDocumentContext()$path 
setwd(dirname(current_path)) 
```

# Linkage Disequilibrium

files: https://www.cog-genomics.org/plink/1.9/formats#bim

## 2. Load the FOXP2.dat file into the R environment. How many individuals and how many SNPs are there in the database? What percentage of the data is missing?

```{r}
# load data
genotypeData <- read.table("FOXP2/FOXP2.dat", header = TRUE)
genotypeData <- genotypeData[, 2: ncol(genotypeData)]
```


## 3. Determine the genotype counts for each SNP, and depict all SNPs simultaeneously in a ternary plot, and comment on your result. For how many variants do you reject Hardy-Weinberg equilibrium using an ordinary chi-square test without continuity correction? (hint: you can read the .bim in R in order to determine the alleles of each SNP, and use function MakeCounts from the HardyWeinberg package to create a matrix of genotype counts). ‘

```{r}
# load data
variantsColnames <- c("chromosome", "SNP id", "position", "basepair coordinate", "Allele 1", "Allele 2")
variantsData <- read.table("FOXP2/FOXP2.bim", header = FALSE, col.names = variantsColnames)
```

## 4. Using the function LD from the genetics package, compute the LD statistic D for the SNPs rs34684677 and rs2894715 of the database. Is there significant association between the alleles of these two SNPs? 

```{r}


```

## 5. Also compute the LD statistic D for the SNPs rs34684677 and rs998302 of the database. Is there significant association between these two SNPs? Is there any reason why rs998302 could have stronger or weaker correlation than rs2894715?

```{r}


```

## 6. Given your previous estimate of D for SNPs rs34684677 and rs2894715, infer the haplotype frequencies. Which haplotype is the most common?

```{r}


```

## 7. Compute the LD statistics R2 for all the marker pairs in this data base, using the LD function of the packages genetics. Be prepared that this make take a few minutes. Also compute an alternative estimate of R2 obtained by using the PLINK program.

For this purpose you should:
* Download and install PLINK 1.90 from https://www.cog-genomics.org/plink2/
* Take care to store the files FOXP2.bim, FOXP2.fam and FOXP2.bed in a directory where PLINK can find them.
* Compute LD estimates with PLINK using plink --bfile FOXP2 --r2 --matrix --out FOXP2 This creates a file with extension FOXP2.ld that contains a matrix with all R2 statistics. Read this file into the R environment. Make a scatter plot for R’s LD estimates against PLINK’s LD estimates. Are they identical or do they at least correlate? What’s the difference between these two estimators? Which estimator would your prefer and why?


```{r}


```

## 8. Compute a distance matrix with the distance in base pairs between all possible pairs of SNPs, using the basepair position of each SNP given in the .bim file. Make a plot of R’s R2 statistics against the distance (expressed as the number of basepairs) between the markers. Comment on your results.

```{r}


```

## 9. Make an LD heatmap of the markers in this database, using the R2 statistic with the LD function. Make another heatmap obtained by filtering out all variants with a MAF below 0.35, and redoing the computations to obtain the R2 statistics in R. Can you explain any differences observed between the two heatmaps?

```{r}


```

## 10. Can you distinguish blocks of correlated markers in the area of the FOXP2 gene? How many blocks do you think that at least seem to exist?

```{r}


```

## 11. Simulate independent SNPs under the assumption of Hardy-Weinberg equilibrium, using R’s sample instruction sample(c("AA","AB","BB"),n,replace=TRUE,prob=c(p\*p,2\*p\*q,q\*q))). Simulate as many SNPs as you have in your database, and take care to match each SNP in your database with a simulated SNP that has the same sample size and allele frequency. Make an LD heatmap of the simulated SNPs, using R2 as your statistic. Compare the results with the LD heatmap of the FOXP2 region. What do you observe? State your conclusions.

```{r}


```

# Haplotype estimation

## 1. 

```{r}
haploData <- read.table("APOE.dat", header = TRUE)
haploData <- haploData[, 2: ncol(haploData)]

```

## 2. 

```{r}


```

## 3. 

```{r}


```

## 4. 

```{r}


```

## 5. 

```{r}


```

## 6. 

```{r}


```

## 7. 

```{r}


```

## 8. 

```{r}


```