---
title: 'Practical 01 SG: Descriptive analysis of genetic markers'
author: "Carlos Moyano & Kleber Reyes"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
editor_options:
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
rm(list = ls(all.names = TRUE))
```

```{r load packages, echo=FALSE, message=FALSE, warning=FALSE}
library(genetics)
library(dplyr)
library(ggplot2)
library(fitdistrplus)
library(HardyWeinberg)
```

# SNP dataset

```{r, include=FALSE}
# load data
filename <- url("http://www-eio.upc.es/~jan/data/bsg/TSICHR22RAW.raw")
italianData <- read.table(filename, header=TRUE)
```

```{r, include=FALSE}
individualData <- italianData[,c(2,5)]
geneticData <- italianData[,7:20655]
rm(italianData)
```

## Questions about SNP dataset

### 3. How many variants are there in this database? What percentage of the data is missing? How many individuals in the database are males and how many are females?



```{r}

variants <- ncol(geneticData); variants # variants in the database
individuals <- nrow(geneticData) 
perc.mis <- 100*sum(is.na(geneticData))/(variants*individuals); perc.mis # 0.1987%

# Let's assume values 1 is male and value 2 is female
male.ind <- length(which(individualData$SEX == 1))
female.ind <- length(which(individualData$SEX == 2))

perc.male <- 100* male.ind / individuals
perc.female <- 100* female.ind / individuals
perc.male; perc.female # 51.96% male - 48.04% female

```

### 4. Calculate the percentage of monomorphic variants (AA or BB). Exclude all monomorphics from the database for all posterior computations of the practical. How many variants do remain in your database?
```{r}

cols <- which(colSums(geneticData == 1, na.rm = TRUE) > 0) # Non monomorphic (contains AB)
variants.poly <-length(cols); variants.poly # 18274 in db
variants.mono <- variants-variants.poly
perc.mono <- 100*variants.mono/variants; perc.mono # 11.50177
geneticData.poly <- geneticData[, cols]

```

### 5. Report the genotype counts and the minor allele count of polymorphism rs8138488_C, and calculate the MAF (Minor Allele Frequency) of this variant.

```{r}

rs8138488_C <- dplyr::recode(geneticData.poly[, "rs8138488_C"], `0`="AA", `1`="AB", `2`="BB")
rs8138488_C.g <- genotype(rs8138488_C,sep="")
rs8138488_C.g.summary <- summary(rs8138488_C.g)
rs8138488_C.g.summary$genotype.freq
rs8138488_C.g.summary$allele.freq
MAF = min(rs8138488_C.g.summary$allele.freq[,"Proportion"]); MAF

```

### 6. Compute the minor allele frequencies (MAF) for all markers, and make a histogram of it. Does the MAF follow a uniform distribution? What percentage of the markers have a MAF below 0.05? And below 0.01? Can you explain the observed pattern?
```{r}
for (i in 1:variants.poly) {
  geneticData.poly[, i] <- dplyr::recode(geneticData.poly[, i], `0`="AA", `1`="AB", `2`="BB")
}
```

```{r}

maf.list <- vector(mode="numeric", length=variants.poly)

ho.list <- vector(mode="numeric", length=variants.poly)
he.list <- vector(mode="numeric", length=variants.poly)

for (i in 1:variants.poly) {
  variant.g <- genotype(geneticData.poly[, i],sep="")
  variant.g.summary <- summary(variant.g)
  
  # MAF
  maf.list[i] <- min(variant.g.summary$allele.freq[,"Proportion"], na.rm = T)
  
  gen.freq <- variant.g.summary$genotype.freq
  # Ho
  if ("A/B" %in% rownames(gen.freq)) {
    ho.list[i] <- gen.freq["A/B",2]
  } else if ("B/A" %in% rownames(gen.freq)) {
    ho.list[i] <- gen.freq["B/A",2]
  }
  
  # He
  sum.p2 <- 0
  if ("A/A" %in% rownames(gen.freq)) {
    sum.p2 <- sum.p2 + gen.freq["A/A",2]
  }
  if ("B/B" %in% rownames(gen.freq)) {
    sum.p2 <- sum.p2 + gen.freq["B/B",2]
  }
  
  he.list[i] <- 1- sum.p2
}

```


```{r}
hist(maf.list, breaks=50, xlab="MAF", main="Histogram of MAF")
descdist(maf.list)
plot(ecdf(maf.list), xlab="MAF", main="ecdf (MAF)")
curve(punif(x, min(maf.list), max(maf.list)), add=TRUE, col="red")

```


As we have seen in the previous plots, the distribution is close to be uniform. 
```{r}

values.under.005 <- length(which(maf.list < 0.05)); values.under.005
values.under.001 <- length(which(maf.list < 0.01)); values.under.001

maf.005 <- 100 * values.under.005 / variants.poly; maf.005
maf.001 <- 100 * values.under.001 / variants.poly; maf.001

```

### 7. Calculate the minor allele frequency for males and for females and present a scatterplot of these variables. What do you observe? Calculate and report their correlation coefficient.
```{r}

geneticData.poly.male <- geneticData.poly[ which( individualData$SEX == 1), ]
geneticData.poly.female <- geneticData.poly[which( individualData$SEX == 2), ]

maf.male <- vector(mode="numeric", length=variants.poly)
maf.female <- vector(mode="numeric", length=variants.poly)

for (i in 1:variants.poly) {
  variant.g <- genotype(geneticData.poly.male[, i],sep="")
  variant.g.summary <- summary(variant.g)
  maf.male[i] = min(variant.g.summary$allele.freq[,"Proportion"], na.rm = T)
  
}

for (i in 1:variants.poly) {
  variant.g <- genotype(geneticData.poly.female[, i],sep="")
  variant.g.summary <- summary(variant.g)
  maf.female[i] = min(variant.g.summary$allele.freq[,"Proportion"], na.rm = T)
  
}

```

```{r}
#maf.value <- c(maf.male, maf.female)
#maf.sex <- c(rep("Male", length(maf.male)), rep("Female", length(maf.female)))
maf.df <- data.frame(maf.male, maf.female)

lgd <- maf.df$maf.sex
ggplot()+geom_point(data=maf.df, aes(x=maf.male, y=maf.female))
```
### 8. Calculate the observed heterozygosity (Ho), and make a histogram of it. What is, theoretically, the range of variation of this statistic?
```{r}
hist(ho.list, breaks = 20)

#sum(geneticData.poly.female=="AA", na.rm=T)
```

### 9. Compute for each marker its expected heterozygosity (He), where the expected heterozygosity for a bi-allelic marker is defined as 1 − E(from i=1 to k) pi^2 , where pi is the frequency of the ith allele. Make a histogram of the expected heterozygosity. What is, theoretically, the range of variation of this statistic? What is the average of He for this database?
```{r}
hist(he.list, breaks = 20)
mean(he.list)
```

```{r}
plot(x=he.list, y=ho.list, xlab="Expected heterozygosity", ylab="Observed heterozygosity", main="Observed vs. Expected heterozygosity")
h.diff = abs(ho.list-he.list)
hist(h.diff, breaks = 3)
```

# STR dataset

## Questions about STR dataset

```{r, include=FALSE}
rm(list=ls())
# Load data
data(NistSTRs)
```

### 2. How many individuals and how many STRs contains the database?
```{r}

X <- NistSTRs
n <- nrow(X) # number of individuals
p <- ncol(X)/2 # number of STRs
n
p

```

There are 361 individuals and 29 STRs.

### 3. Write a function that determines the number of alleles for a STR. Determine the number of alleles for each STR in the database. Compute basic descriptive statistics of the number of alleles (mean,
standard deviation, median, minimum, maximum).
```{r}

# Function that determines the number of alleles for a STR.
n.alleles <- function(X, str.index) {
  allele.1 <- as.list(X[,str.index])
  allele.2 <- as.list(X[,(str.index+1)])
  return(length(table(unlist(c(allele.1, allele.2))))) # number of alleles
}

n.alleles.per.str.list <- list()
str.index <- 1
for (str.num in 1:p) {
  n.alleles.per.str.list  <- append(n.alleles.per.str.list, n.alleles(X, str.index))
  str.index <- str.index + 2
}
n.alleles.per.str <- unlist(n.alleles.per.str.list)

# Basic descriptive statistics of the number of alleles
mean(n.alleles.per.str)
sd(n.alleles.per.str)
median(n.alleles.per.str)
max(n.alleles.per.str)
min(n.alleles.per.str)

```

### 4. Make a table with the number of STRs for a given number of alleles and present a barplot of the number STRs in each category. What is the most common number of alleles for an STR?
```{r, echo=FALSE}

barplot(table(n.alleles.per.str), xlab="Number of alleles", ylab="Number of STRs")

```

The most common number of alleles for an STR is 8.

### 5. Compute the expected heterozygosity for each STR. Make a histogram of the expected heterozygosity over all STRS. Compute the average expected heterozygosity over all STRs.
```{r, out.width = "80%"}

exp.heter <- function(X, str.index) {
  allele.1 <- as.list(X[,str.index])
  allele.2 <- as.list(X[,(str.index+1)])
  t <- table(unlist(c(allele.1, allele.2)))
  sum.t <- sum(unname(t)) # we sum the counts
  exp.heter <- round(1 - sum(sapply(unname(t), function(x) (x / sum.t)^2 )), 3)
  return(exp.heter) # expected heterozygosity formula
}

exp.heter.per.str.list <- list()
str.index <- 1
for (str.num in 1:p) {
  exp.heter.per.str.list  <- append(exp.heter.per.str.list, exp.heter(X, str.index))
  str.index <- str.index + 2
}
exp.heter.per.str <- unlist(exp.heter.per.str.list)

hist(exp.heter.per.str, xlab="Expected heterozygosity", main="Histogram of the expected heterozygosity")

round(mean(exp.heter.per.str), 3) # average expected heterozygosity over all STRs

```

### 6. Calculate also the observed heterozygosity for each STR. Plot observed against expected heterozygosity, using all STRs. What do you observe? (Ho = fAB)
```{r}

obs.heter <- function(X, str.index) {
  
  allele.1 <- X[,str.index]
  allele.2 <- X[,str.index+1]
  allele.1n <- pmin(allele.1,allele.2)
  allele.2n <- pmax(allele.1,allele.2)
  
  index_different <- allele.1n != allele.2n
  
  individuals_heter <- paste(allele.1n[index_different], allele.2n[index_different],sep="/")
  individuals_heter
  
  individuals <- paste(allele.1n, allele.2n,sep="/")
  g.counts.sum <- sum(table(individuals))
  
  g.heter.counts.sum <- sum(table(individuals_heter))
  g.heter.counts.sum
  
  Ho <- round(g.heter.counts.sum / g.counts.sum, 3)
  
  return(Ho)
}

obs.heter.per.str.list <- list()
str.index <- 1
for (str.num in 1:p) {
  obs.heter.per.str.list  <- append(obs.heter.per.str.list, obs.heter(X, str.index))
  str.index <- str.index + 2
}
obs.heter.per.str <- unlist(obs.heter.per.str.list)

```

```{r echo=FALSE, out.width = "80%"}

# Plot observed against expected heterozygosity, using all STRs
plot(x=exp.heter.per.str, y=obs.heter.per.str, xlab="Expected heterozygosity", ylab="Observed heterozygosity", main="Observed vs. Expected heterozygosity")

```

### 7. Compare, overall, the results you obtained for the SNP database with those you obtained for the STR database. What differences do you observe between these two types of genetic markers?
