---
title: 'Practical 01 SG: Descriptive analysis of genetic markers'
author: "Carlos Moyano & Kleber Reyes"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup environment

## Clear environment
```{r clear}
rm(list = ls(all.names = TRUE))
```

## Load packages
```{r load packages, message=FALSE}
library(genetics)
```

# SNP dataset

## Load data
```{r clear}
filename <- url("http://www-eio.upc.es/~jan/data/bsg/TSICHR22RAW.raw")
italianData <- read.table(filename, header=TRUE)
ls()
```

## Load data
```{r clear}
individualData <- italianData[,c(2,5)]
geneticData <- italianData[,7:20655]
```

Notice that the genetic variants are identifed by an “rs” identifier.
The genetic data is coded in the (0, 1, 2) format with 0=AA, 1=AB, 2=BB.

## Questions

### How many variants are there in this database? What percentage of the data is missing? How many individuals in the database are males and how many are females?
```{r}

```

### Calculate the percentage of monomorphic variants. Exclude all monomorphics from the database for all posterior computations of the practical. How many variants do remain in your database?
```{r}

```

### Report the genotype counts and the minor allele count of polymorphism rs8138488_C, and calculate the MAF of this variant.
```{r}

```

### Compute the minor allele frequencies (MAF) for all markers, and make a histogram of it. Does the MAF follow a uniform distribution? What percentage of the markers have a MAF below 0.05? And below 0.01? Can you explain the observed pattern?
```{r}

```

### Calculate the minor allele frequency for males and for females and present a scatterplot of these variables. What do you observe? Calculate and report their correlation coefficient.
```{r}

```

### Calculate the observed heterozygosity (Ho), and make a histogram of it. What is, theoretically, the range of variation of this statistic?
```{r}

```

### Compute for each marker its expected heterozygosity (He), where the expected heterozygosity for a bi-allelic marker is defined as 1 − E(from i=1 to k) pi^2 , where pi is the frequency of the ith allele. Make a histogram of the expected heterozygosity. What is, theoretically, the range of variation of this statistic? What is the average of He for this database?
```{r}

```